---
title: "Campfire Response Times"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Comparing Response Times to the Camp Fire of the Public and Local Officials | Olivia Plazanet, Madeline Hoshko and Taylor Hinchliff


## Introduction and Background


## Introduction

- <font size="3">For our topic, we chose to examine the Tweet response times of official accounts such as CalFire, the Butte County Sheriff, and others during the Camp Fire, the deadliest and most destructive wildfire in California history and the most expensive natural disaster in the world in 2018 in terms of insured losses (USA Today, 2019).</font>
- <font size="3">Investigating these kinds of response times  is important, because of the need of accurate, timely warnings and other information should a disaster, natural or otherwise, occur and threaten the lives and livelihoods of people.</font>
- <font size="3">In the case of previous popular means of communication and warning such as radio, local television, and even direct phone calls, we can no longer rely on these methods reaching every citizen in order to warn them about an impending disaster. Instead, other methods are needed that are more widely used.</font>
- <font size="3">But unfortunately, a tool is only as useful as the person who uses it. If used incorrectly or inappropriately, it could lead to further danger by misleading or misinforming the public. That is why quick, accurate, and widespread posts with pertinent information can be useful in times of emergency.</font>
- <font size="3">Algorithms, or set instructions used to perform a simple task in programming, can be used in Twitter to see “trending” or popular terms and hashtags, which can be incredibly useful in times of a crisis if they are used to identify where and what is happening.</font>
- <font size="3">That is why it is also very important to examine the times it takes for officials to issue official information and warnings to the public, such as evacuation notices. And for that reason we have chosen to study the response times of official accounts during the Camp Fire on Twitter, one of the most popular social media platforms in the United States.</font>

## Why Use Twitter?
- <font size="2.75">Social network Twitter is particularly popular in the United States, where as of April 2020, the microblogging service had audience reach of 64.2 million users.</font>
- <font size="2.75">In the Nielsen ratings for 2018-19, the broadcast networks averaged 28.5 million viewers in primetime, a decline of 7.3 million viewers (20 percent) since 2014-15, so television is becoming less helpful in distributing timely information.</font>
- <font size="2.75">Radio is also a poor choice for the distribution of time-sensitive announcements, The Infinite Dial research for the past 12 years shows how the radio has been declining in the home. “In 2008 only 4% of Americans reported having no radio in their home, as compared to 32% today (Radio Ink, 2020).</font>
- <font size="2.75">Not only is social media used by a wide variety of demographics, but it is also used for a significant portion of those citizens’ days and is starting to replace time spent with traditional media. For local governments that have primarily focused their citizen communication strategies around news distribution through traditional media outlets, we now know that social media serves as a more direct and more impactful way to communicate with citizens (EfficientGov, 2018).
News now usually appears on social media first.</font>
  - <font size="2">Social media crosses geographic boundaries more quickly.</font>
  - <font size="2">Social channels are informed by multiple individuals firsthand, often through unofficial sources, but sometimes those who are direct witnesses to events.</font>
  - <font size="2">Social media is commented upon by active social media users who further share and distribute content.</font>
- <font size="2.75">Journalists watch social media to identify and report breaking news stories. According to Insignia, 77 percent of journalists say social media is important to them for learning about potential stories more quickly, and the same percentage say social media is important in reporting stories more quickly (Elliott 2018).</font>

## Background and Lit Review 

- <font size="5">Disaster Response and Twitter</font>
  - <font size="4">Why is Twitter data important during disasters?</font>
    - <font size="3">Twitter gives everyone a “real-time view” of disasters, allowing them to get information quickly from official sources regarding their safety (Ford, 2018).</font>
  - <font size="4">"Given the volume of data produced, manually managing this process in the immediate aftermath of a crisis is not always practical. There is also often a need for unique updates related to particular topics within and across organizations” (Ford, 2018).</font>
  - <font size="4">Some groups are coming up with programs specifically tailored towards combing Twitter during disasters for noun-verb pairs that might come up during disasters like “bridge collapsed” and “person trapped”. This could help official response times in future disasters (Ford, 2018).</font>
  
## Background and Lit Review 

- <font size="5">Potential Flaws in Phone-Based Warning Systems</font>
  - <font size="4">“Call failure reports released by the Butte County Sheriff’s Office at the request of The Times show that the first evacuation orders requested by firefighters at the scene of the massive blaze frequently failed to connect. About as many calls went to voicemail as were answered by a live person (St. John and Serna, 2018).”</font>
  - <font size="4">If officials are not able to make timely announcements and easily warn the public of a looming disaster, this presents a real problem within our system.</font>
  - <font size="4">In the case of the Camp Fire, the police chief of Paradise said “there was no time for a citywide evacuation order — the city’s own system went down in the midst of a partial order” (St. John and Serna, 2018). This shows a need for a wide array of emergency notification methods, and with California being the state with the largest amount of Twitter users, the platform seems perfect to fit the role of one of these notification methods (Stirtz, 2020).</font>

## Background and Lit Review 

- <font size="5">Refiguration of disaster social relations</font>
  - <font size="4">The proliferation of new communication technologies is replacing traditional media when it comes to the flow of information, there is a transformation in the visibility of disasters (Cottle 2014).</font>
  - <font size="4">Plays a more progressive part in disasters: connected individuals are becoming more central in emergency responses in humanitarian crisis situations (Cottle 2014).</font>
  - <font size="4">“Challenging the top-down, elite dominated communications” (Cottle 2014)
There is a civilian surge of information as well as worldwide engagement because of fast-flowing information, collaboration between citizens, disaster relief agencies, volunteers, and public figures (Cottle 2014).</font>

## Background and Lit Review 

- <font size="5">Limits to crisis data</font>
  - <font size="4">Once a disaster occurs, data collection starts. However, using social media platforms such as Twitter can create analytical and ethical oversights. Not everybody has access to the internet, where the disaster occurred there can be power outages, etc (Crawford and Finn 2015).</font>
  - <font size="4">Before social media, traditional media was state-controlled, and governments were used to it, but that changed in more recent years (Crawford and Finn 2015).</font>
  - <font size="4">The Twitter algorithm can influence the way tweets are created or how they are retweeted; “bots” can also impede data-collection (Crawford and Finn 2015).</font>
  - <font size="4">People cannot control how their data is used, which can pose a privacy issue. It is important to examine underlying assumptions when using social or mobile data (Crawford and Finn 2015).</font>

## Research Problem and Topic Area
- <font size="5">How timely were the twitter accounts run by public officials and emergency services when it came to disseminating important and crucial information to the public during the Camp Fire? We will create a timeline of events to illustrate this problem.</font>
- <font size="5">While there is a large amount of research that has been done regarding the usefulness of social media and its innovative uses in emergency management, the majority of the case studies have revolved around incidents such as earthquakes. In a study conducted by the US Department of Homeland Security in 2013, out of the six cases that were examined only one of them pertained to wildfires. One of the goals of our research is to examine the effectiveness of social media in the role of emergency management and expand upon its usefulness in cases of emergencies, in this case specifically the Camp Fire.</font>

## Data Introduction
- <font size="3">74,099  tweets were collected by the Spring 2020 Advanced Data Science class at CSU, Chico. Tweets were collected using the rtweet (Kearney MW 2019) package using R 3.6.2 (R Core Team) beginning the morning of November 11, 2018.</font>
- <font size="3">A timeline of events from the morning of November 8, 2018 was derived from the Chico Enterprise-Record. (Epley)</font>
- <font size="3">Each record in the data represents an individual tweet and contains all the information regarding the tweet in terms of who posted it, the times its been retweeted, and more. We are looking specifically into the time of day each tweet was posted to look into the response times of organizations in charge of monitoring the situation.</font>
- <font size="3">During the data collection the only data that needed to be discarded were results that turned up not belonging to our query as they were not in English. Otherwise, our query that filtered tweets to include one or more of the following: #CampFire, #campfire, #paradise, #Paradise or @CALFIRE_ButteCo" during the correct time frame.</font>
- <font size="3">The reliability of data collection is important to our research question as we are interested in few accounts and want to ensure we have the most information as possible and that it is accurate since we are concerned with such a small amount of data compared to the entirety of the dataset.</font>
- <font size="3">We are observing tweets regarding seven public figures’ and organizations’ twitter accounts that were instrumental in dispersing information regarding the Camp Fire, specifically: @Cal_Fire, @CalFire_ButteCo, @ButteSherriff, @ChicoFD, @CountyOfButte, @Paradise_CA, and @ChicoPolice.</font>
- <font size="3">The data is defined to be a media source when it belongs to one of the previous accounts, while all others are classified as public.</font>


## Methods
- <font size="5">Variable Creation: To successfully compare the response time of the following accounts: @Cal_Fire, @CalFire_ButteCo, @ButteSherriff, @ChicoFD, @CountyOfButte, @Paradise_CA, and @ChicoPolice to the real life timing of events we referenced Chico Enterprise-Record’s timeline of the fire. (Epley) We populated a data frame with each event and time as specified by Chico Enterprise-Record.</font>
- <font size="5">Data Analysis: We analyzed the specific timing of emergency services and government entities’ tweets regarding pertinent information in relationship to the real life events through a timeline that incorporates both the twitter data and the derived timeline data in one merged timeline plot. We additionally created a timeline documenting the comparison between times of general public tweets and tweets that came from our selected local media sources stated above. Both timelines show the event description or tweet content along with the time the event or tweet occured.</font>

## Description of variables being used
- <font size="5">The data for the Camp Fire that we were most interested in involved initial response times, so we filtered down the dataset to only include tweets that occurred during the first six hours from the fire starting.</font>
- <font size="5">The variables used throughout include:</font>
  - <font size="5">Tweet content (text)</font>
  - <font size="5">User’s Twitter Handle (screen_name)</font>
  - <font size="5">Whether the tweet was a retweet (is_retweet)</font>
  - <font size="5">Time the tweet was created (created_at_pst)</font>
  - <font size="5">The events derived from Chico Enterprise-Record’s timeline (Epley)</font>

## Results

## Univariate Description of Explanatory Measure
```{r, include = FALSE}
pkgs = c("stringr", "dplyr", "tidyverse", "readr", "ggplot2", "knitr", "curl", "gridExtra")
for (pkg in pkgs){
  if (!require(pkg, character.only = T)){
    install.packages(pkg)
    library(pkg)
  }
}
purl("ProjectEDA.Rmd", output="part1.r")
source("part1.r")
```
<div style="float: left; width: 40%;">
<font size="5">While the original dataset includes 74,099 tweets, we are only concerned with the tweets that occur within the first 6 hours of the fire starting. This ends up being just 446 tweets, which amounts to 0.60% of the overall data. Of these we have 425 (95.51%) classified as public tweets and 21 (4.71%) classified as media tweets (tweets from our local media accounts). Our explanatory variable is defined as the local media accounts. </font>
</div>

<div style="float: right; width:60%;">
```{r, echo=FALSE, fig.width=6.6}
graph1
```
</div>

## Univariate Description of Response Measure 
<div style="float: left; width: 40%;">
<font size="5">Our response measure is the time of the tweets as we are interested in the difference in time between real life events and tweets. Here we show the overall timeline of the first six hours of tweets that we will be referencing throughout. </font>
<br>
<br>
```{r, echo=FALSE, fig.width=5}
pander(overall_time)
```
</div>

<div style="float: right; width:60%;">
```{r, echo=FALSE, fig.width=6.6}
graph2
```
</div>

## Bivariate Description of Response 
<div style="float: left; width: 40%;">
<font size="5">The response times and amount of responses for each account can be shown here to display the accounts that were most involved within the first hours of the fire. 
The media’s first response to the fire on Twitter appears at 06:51:47	by CAL FIRE Butte County</font>
<br>
<br>
```{r, echo=FALSE, fig.width=5}
pander(media_time)
```
</div>

<div style="float: right; width: 60%;">
```{r, echo=FALSE, fig.width=6.8}
graph3
```
</div>

## Public Tweets
<div style="float: left; width: 40%;">
<font size="5">Here we can see the activity of public accounts to see the frequency of tweets. What’s important to note is the vast number of retweets, showing the number of tweets that are typically dispersing the media’s tweets to reach a wider audience. </font>

<font size="5">The first tweet by a public user closely followed CAL FIRE Butte County’s at 06:54:55</font>
</div>

<div style="float: left; width: 60%;">
```{r, echo = FALSE, fig.width=6.8}
graph4
```
</div>

## The Media Response vs Real-Time Events
<font size="4.5">In the following timeline each dot signifies a tweet or event to show tweet times in comparison to the real event times. The majority of the tweets shown are evacuation warnings and orders. </font>

```{r pressure, echo=FALSE}
plot1
```

## The Media Response vs Public Response
<font size="4.5">After eliminating retweets the following timeline compares the tweet frequency of the public vs the media for the morning up until 10:00 AM. </font>

```{r, echo=FALSE}
plot2
```

## Discussion/Conclusion

## References
<font size="4">Cottle, Simon. (2014). Rethinking Media and Disasters in a Global Age: What's Changed and Why It Matters. Media, War & Conflict. 7. 3-22.10.1177/1750635213513229. https://www.researchgate.net/publication/270633894_Rethinking_Media_and_Disasters_in_a_Global_Age_What's_Changed_and_Why_It_Matters </font>

<font size="4">Elliott, Jennifer. “8 Best Practices for Emergency Communications on Social Media.” EfficientGov, 19 July 2018, 
https://www.efficientgov.com/community-engagement/articles/8-best-practices-for-emergency-communications-on-social-media-vwZS7OO5eoblrs3G/. </font>

<font size="4">Epley, Robin (November 8, 2019). “Timeline: Breaking down Nov. 8 - the Day the Camp Fire Sparked.” Chico Enterprise-Record, Chico Enterprise-Record 
https://www.chicoer.com/2019/11/07/timeline-breaking-down-nov-8-the-day-the-camp-fire-sparked/. </font>

<font size="4">Ford, Jordan. “Improving Disaster Response through Twitter Data.” Penn State University, 2018, 
https://news.psu.edu/story/527730/2018/07/10/research/improving-disaster-response-through-twitter-data. </font>

<font size="4">Kearney MW (2019). “rtweet: Collecting and analyzing Twitter data.” Journal of Open Source Software, 4(42), 1829. Doi: 10.21105/joss.01829, R package version 0.7.0, https://joss.theoj.org/papers/10.21105/joss.01829. </font>
 



## References cont. 

<font size="4">R Core Team (2019). R: A Language and Environment for Statistical Computing. R Foundation for Statistical Computing, Vienna, Austria. URL https://www.R-project.org. </font>

<font size="4">Radio Ink (March 3, 2020). “The Decline of the Home Radio”.  https://radioink.com/2020/03/03/the-decline-of-the-home-radio/ </font>

<font size="4">Rice, Doyle (January 8, 2019). “USA had world’s 3 costliest natural disasters in 2018, and Camp Fire was the worst”. USA Today. </font>

<font size="4">St. John, Paige, and Joseph Serna. “Camp Fire Evacuation Warnings Failed to Reach More than a Third of Residents Meant to Receive Calls.” Los Angeles Times, 1 Dec. 2018, 
https://www.latimes.com/local/california/la-me-ln-paradise-evacuation-warnings-20181130-story.html. </font>

<font size="4">Stirtz, Kevin. “Twitter Ranking: Which States Twitter the Most?” All Business, Dun & Bradstreet, 14 May 2009, https://www.allbusiness.com/twitter-ranking-which-states-twitter-the-most-12329567-1.html. </font>
